# Introduction

<!-- (C) 2020 Marek Gagolewski, https://www.gagolewski.com -->


<!-- TODO: citations


-->



## What is Machine Learning?




An *algorithm* is a well-defined sequence of instructions that,
for a given sequence of input arguments,
yields some desired output.
In other words, it is a specific recipe for what we call
a *function* in mathematics.
Unfortunately, algorithm development is a tedious task. We need to be
super-precise about covering all the possible scenarios and modelling them
accurately.


In *machine learning* (ML), we build and study computer algorithms
that make *predictions* or *decisions* but which are not
manually programmed. Moreover, some algorithms might also be able
to *discover* new interesting facts about a problem instance at hand.

*Learning*, however, needs some material based upon which
new knowledge is to be acquired.
In other words, we need *data*.


### Data Sources

Information can come from various sources, e.g., physical sensors,
files, databases, (pseudo)random number generators
and it can be of different forms, e.g., vectors, matrices
and other tensors, graphs, audio/video streams, text etc.
With the advent of the internet era, data have become ubiquitous.

{ BEGIN exercise }
Think of how much information you consume and generate when you
interact with your social media or news feeds every day.
{ END exercise }



Here are some application domains where machine learning has already proven
itself very useful:

* Financial services (banking, insurance, investment funds)
* Oil and gas and other energy (solar, mining)
* Real estate
* Pharmaceuticals
* Advertising
* Transportation
* Retail
* Healthcare
* Food production

{ BEGIN exercise }
Think of different ways in which
these sectors could benefit from ML solutions.
{ END exercise }

To be frank, the above list was generated by duckduckgoing
the "biggest industries" query. That was a very easy task;
ML is already everywhere.
Basically, wherever we have data and there is a need to improve
some processes or discover new aspects about a problem domain,
there is a place for ML solutions.

Of course, it's not all about business revenue (luckily).
We can do a lot of great work for greater good;
with the increased availability of open data,
everyone can be a reporter, an engaged citizen that seeks for truth.
There are NGOs.
Finally, there are researchers (remember that in many countries
still the  main role of a university is to spread knowledge and not make money!)
that need these methods to make new discoveries, e.g., in psychology,
sociology, agriculture, engineering, biotechnology, pharmacy, medicine, genetics,
you name it.

<!-- * Life science and medical data analysis -->
<!-- * Scientific simulations -->
<!-- * Genomics -->



### Main Types of Machine Learning Problems

Machine Learning problems include, but are not limited to:

- *Supervised learning* -- for every input point (e.g., a photo)
there is an associated desired output (e.g., whether it depicts a crosswalk
or how many cars can be seen on it);
in other words, there is a "teacher" ready to instruct us
about what is expected the expected outcome in a given context


- *Semi-supervised learning* -- some inputs are labelled, other ones
are not (definitely a cheaper scenario)



- *Unsupervised learning* -- inputs are unlabelled, the aim is to discover
the underlying structure in the data (e.g., automatically group customers
w.r.t. common behavioural patterns);
think about how much you learnt by just interacting with your
environment in a random fashion



- *Reinforcement learning* -- learn to act based on a
feedback given after the actual decision was made
such as "that was nice! well done! look at you go!"
(e.g., learn to play The Witcher 7 by testing different hypotheses
what to do to survive as long as possible)





In the course of this book,
we're going to take a deep dive into what AI really is, what
are its limitations but also how we can use it for improving different processes.


outputs can be interpreted by a human being and used (critically) for whatever  purpose
(e.g., a GP for diagnosing illnesses)
* outputs can be provided to other algorithms to take immediate actions
(e.g., buy those market shares immediately!)


## Supervised Learning

### Formalism




Let $\mathbf{X}=\{\mathfrak{X}_1,\dots,\mathfrak{X}_n\}$
be an input sample ("a database")
that consists of $n$ objects.



Most often we assume that each object $\mathfrak{X}_i$
is represented using $p$ numbers for some $p$.

We denote this fact as $\mathfrak{X}_i\in \mathbb{R}^p$
(it is *a $p$-dimensional real vector* or
*a sequence of $p$ numbers* or
*a point in a $p$-dimensional real space*
or *an element of a real $p$-space* etc.).

If we have "complex" objects on input,
we can always try representing them as **feature vectors** (e.g.,
come up with numeric attributes that best describe them in a task at hand).

{ BEGIN exercise }
Consider the following problems:

1. How would you represent a patient in a clinic?

2. How would you represent a car in an insurance company's database?

3. How would you represent a student in an university?
{ END exercise }





Of course, our setting is *abstract* in the sense that
there might be different realities *hidden* behind these symbols.

This is what maths is for -- creating *abstractions* or *models*
of complex entities/phenomena so that they can be much more easily manipulated
or understood.
This is very powerful -- spend a moment
contemplating how many real-world situations fit into our framework.



This also includes image/video data, e.g., a 1920×1080 pixel image
can be "unwound" to a "flat" vector of length 2,073,600.

(\*) There are some algorithms such as Multidimensional Scaling,
Locally Linear Embedding, IsoMap etc.
that can do that automagically.

. . .



In cases such as this we say that we deal with *structured (tabular) data*\
-- $\mathbf{X}$ can be written as an ($n\times p$)-matrix:
\[
\mathbf{X}=
\left[
\begin{array}{cccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} \\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} \\
\end{array}
\right]
\]
Mathematically, we denote this as $\mathbf{X}\in\mathbb{R}^{n\times d}$.





Remark.

: Structured data == think: Excel/Calc spreadsheets, SQL tables etc.








For an example, consider the famous Fisher's Iris flower dataset,
see `?iris` in R
and https://en.wikipedia.org/wiki/Iris_flower_data_set.

```{r show_iris}
X <- iris[1:6, 1:4] # first 6 rows and 4 columns
X         # or: print(X)
dim(X)    # gives n and p
dim(iris) # for the full dataset
```




$x_{i,j}\in\mathbb{R}$
represents the $j$-th feature of the $i$-th observation,
$j=1,\dots,p$, $i=1,\dots,n$.

For instance:

```{r show_iris_elem}
X[3, 2] # 3rd row, 2nd column
```





The third observation (data point, row in $\mathbf{X}$)
consists of items $(x_{3,1}, \dots, x_{3,p})$ that can be extracted by calling:

```{r show_iris_row}
X[3,]
as.numeric(X[3,]) # drops names
```

```{r show_iris_row_length}
length(X[3,])
```




Moreover, the second feature/variable/column
is comprised of
$(x_{1,2}, x_{2,2}, \dots, x_{n,2})$:

```{r show_iris_col}
X[,2]
length(X[,2])
```





We will sometimes use the following notation to emphasise that
the $\mathbf{X}$ matrix  consists of $n$ rows
or $p$ columns:

\[
\mathbf{X}=\left[
\begin{array}{c}
\mathbf{x}_{1,\cdot} \\
\mathbf{x}_{2,\cdot} \\
\vdots\\
\mathbf{x}_{n,\cdot} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
\mathbf{x}_{\cdot,1} &
\mathbf{x}_{\cdot,2} &
\cdots &
\mathbf{x}_{\cdot,p} \\
\end{array}
\right].
\]




Here, $\mathbf{x}_{i,\cdot}$ is a *row vector* of length $p$,
i.e., a $(1\times p)$-matrix:

\[
\mathbf{x}_{i,\cdot} = \left[
\begin{array}{cccc}
x_{i,1} &
x_{i,2} &
\cdots &
x_{i,p} \\
\end{array}
\right].
\]

Moreover, $\mathbf{x}_{\cdot,j}$ is a *column vector* of length $n$,
i.e., an $(n\times 1)$-matrix:

\[
\mathbf{x}_{\cdot,j} = \left[
\begin{array}{cccc}
x_{1,j} &
x_{2,j} &
\cdots &
x_{n,j} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{x}_{1,j} \\
{x}_{2,j} \\
\vdots\\
{x}_{n,j} \\
\end{array}
\right],
\]

where $\cdot^T$ denotes the *transpose* of a given matrix --
think of this as a kind of rotation; it allows us to introduce a set of
"vertically stacked" objects using a single inline formula.



### Desired Outputs




In supervised learning,
apart from the inputs we are also given the corresponding
reference/desired outputs.

The aim of supervised learning is to try to create an "algorithm" that,
given an input point, generates an output that is as *close* as possible
to the desired one. The given data sample will be used to "train" this "model".

Usually the reference outputs are encoded as individual  numbers (scalars)
or textual labels.

In other words, with each input $\mathbf{x}_{i,\cdot}$ we associate
the desired output $y_i$:

```{r show_iris_sample,echo=-1}
set.seed(123)
# in iris, iris[, 5] gives the Ys
iris[sample(nrow(iris), 3), ]  # three random rows
```





Hence, our dataset is $[\mathbf{X}\ \mathbf{y}]$ --
each object is represented as a row vector
$[\mathbf{x}_{i,\cdot}\ y_i]$, $i=1,\dots,n$:

\[
[\mathbf{X}\ \mathbf{y}]=
\left[
\begin{array}{ccccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} & y_1\\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} & y_2\\
\vdots & \vdots & \ddots & \vdots    & \vdots\\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} & y_n\\
\end{array}
\right],
\]

where:

\[
\mathbf{y} = \left[
\begin{array}{cccc}
y_{1} &
y_{2} &
\cdots &
y_{n} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{y}_{1} \\
{y}_{2} \\
\vdots\\
{y}_{n} \\
\end{array}
\right].
\]



### Types of Supervised Learning Problems




Depending on the type of the elements in $\mathbf{y}$
(the domain of $\mathbf{y}$),
supervised learning problems are usually
classified as:

- **regression** -- each $y_i$ is a real number

    e.g., $y_i=$ future market stock price
    with $\mathbf{x}_{i,\cdot}=$ prices from $p$ previous days

- **classification** -- each $y_i$ is a discrete label

    e.g., $y_i=$ healthy (0) or ill (1)
    with $\mathbf{x}_{i,\cdot}=$ a patient's health record


- **ordinal regression** (a.k.a. ordinal classification) -- each $y_i$ is a rank

    e.g., $y_i=$ rating of a product on the scale 1--5
    with $\mathbf{x}_{i,\cdot}=$ ratings of $p$ most similar products





{ BEGIN exercise }
Example Problems -- Discussion:

Which of the following are instances of classification problems? Which of them are regression tasks?

What kind of data should you gather in order to tackle them?

- Detect email spam
- Predict a market stock price
- Predict the likeability of a new ad
- Assess credit risk
- Detect tumour tissues in medical images
- Predict time-to-recovery of cancer patients
- Recognise smiling faces on photographs
- Detect unattended luggage in airport security camera footage
- Turn on emergency braking to avoid a collision with pedestrians
{ END exercise }




A single dataset can become an instance of many different ML problems.

Examples -- the `wine_quality` dataset:

```{r wines_load,echo=-1}
set.seed(123)
wine_quality <- read.csv("datasets/wine_quality_all.csv",
    comment.char="#")
wine_quality[1,]
```


`alcohol` is a numeric (quantitative) variable (see Figure {@fig:wines_regression} for a histogram depicting its empirical distribution):


```{r wines_regression,fig.cap="Quantitative (numeric) outputs lead to regression problems"}
summary(wine_quality$alcohol) # continuous variable
hist(wine_quality$alcohol, main="", col="white"); box()
```

`color` is a quantitative variable with two possible outcomes (see Figure {@fig:wines_binary} for a bar plot):



```{r wines_binary,fig.cap="Quantitative outputs lead to classification tasks"}
table(wine_quality$color) # binary variable
barplot(table(wine_quality$color), col="white", ylim=c(0, 6000))
```


Moreover, `response` is an ordinal variable, representing
a wine's rating as assigned by a wine expert
(see  Figure {@fig:wines_ordinal} for a barplot).
Note that although the ranks are represented with numbers,
we they are not continuous variables. Moreover,
these ranks are something more than just labels -- they are linearly
ordered, we know what's the smallest rank and whats the greatest one.

```{r wines_ordinal,fig.cap="Ordinal variables constitute ordinal regression tasks"}
table(wine_quality$response) # ordinal variable
barplot(table(wine_quality$response), col="white", ylim=c(0, 3000))
```


