# Introduction

<!-- (C) 2020 Marek Gagolewski, https://www.gagolewski.com -->


<!-- TODO: citations


-->



<!-- TODO



We need more than just ML:

Mathematical modelling, differential equations, discrete maths

Statistics for small samples but not needed for big data




It's not all in the data, real life processes are not static

Machine learning models are inherently static, they describe today. Tomorrow may be different, especially if people start to trick the system. Think spam detection , spammers will change their behaviour as new detection algorithms arise

People are smarter than your think, they will challenge the system, by choosing a quality metric you change the direction, but can you really predict what other factors gonna be like when we get there?


Examples : covid prediction not taking into account country policies and other things going on

Garmin vo2max estimation and the risk of going mad/overtraining


-->


<!-- TODO


# What is AI?

**Artificial Intelligence** (AI) aims to construct algorithms and devices
to *make decisions that maximise their chance of successfully achieving
their goals*.



We build AI systems to **solve various problems** under
the assumption of **availability of data**.



Crucial features of AI systems include the ability to:

* perceive the environment

    * input data can come from various sources, e.g., physical sensors,
    files, databases, (pseudo)random number generators
    * data can be of different forms, e.g., vectors, matrices
    and other tensors, graphs, audio/video streams, text

* learn from past experience

    * information on how to correctly behave (what is expected of the system)
    in certain situations might be available
    * sometimes feedback can be given after making a decision ("that was nice!
    well done! look at you go!")
    * some systems can adapt based on current experience as well
    (learn to modify its behaviour *on-line*)

* make decisions, predictions, explanations

    * outputs can be interpreted by a human being and used (critically) for whatever  purpose
    (e.g., a GP for diagnosing illnesses)
    * outputs can be provided to other algorithms to take immediate actions
    (e.g., buy those market shares immediately!)



# Common Applications


Application domains of AI include, but are not limited to:


* Life science and medical data analysis
* Financial services (banking, insurance, investment funds)
* Oil and gas and other energy (solar, mining)
* Real estate
* Pharmaceuticals
* Scientific simulations
* Genomics
* Advertising
* Transportation
* Retail
* Healthcare
* Food production

> Discussion (exercise):
Think of different ways in which
these sectors could benefit from AI solutions.

Basically, wherever we have data and there is a need to improve
things, there is a place for AI solutions.

In the course of this book,
we're going to take a deep dive into what AI really is, what
are its limitations but also how we can use it for improving different processes.

 *AI is a tool*.
Yes, AI is really useful, but it's not a remedy for all our problems.


Behind every success story related to  AI, there are always people involved. Sometimes large
teams of:

* those who deal with hardware, data storage, high performance computing
* those who wrangle data, fit models to data and develop prototypes
* those who are responsible for making the algorithms
work well in production environments
* those who are responsible for providing you
with software/algorithms/models that enable everyone to perform the above,
e.g., languages such as R and Python, libraries, packages

You can call them data engineers, data analysts, machine learning engineers,
data scientists, whatever.

We would like to appreciate the role of open source software.
Most started as non-for-profit and it's beautiful that some of them
continue this way.
Not everything is for sale. Not everything has its price.
Let's not forget it.

We can do a lot of great work for greater good,
with the increased availability of open data,
everyone can be a reporter, an engaged citizen that seeks for truth.
There are NGOs.
Finally, there are researchers (remember that in many countries
still the university's main role is to spread knowledge and not make money!)
that need these methods to make new discoveries, e.g., in psychology,
sociology, agriculture, engineering, biotechnology, pharmacy, medicine, you name it.
-->

## What is Machine Learning?




An **algorithm** is a well-defined sequence of instructions that,
for a given sequence of input arguments,
yields some desired output.

In other words, it is a specific recipe for a **function**.

Developing algorithms is a tedious task.





In **machine learning**, we build and study computer algorithms
that make *predictions* or *decisions* but which are not
manually programmed.

**Learning** needs some material based upon which new knowledge is to be acquired.

In other words, we need **data**.



### Main Types of Machine Learning Problems




Machine Learning Problems include, but are not limited to:

- **Supervised learning** -- for every input point (e.g., a photo)
there is an associated desired output (e.g., whether it depicts a crosswalk
or how many cars can be seen on it)

- **Unsupervised learning** -- inputs are unlabelled, the aim is to discover
the underlying structure in the data (e.g., automatically group customers
w.r.t. common behavioural patterns)

- **Semi-supervised learning** -- some inputs are labelled, the others
are not (definitely a cheaper scenario)


- **Reinforcement learning** -- learn to act based on a
feedback given after the actual decision was made
(e.g., learn to play The Witcher 7 by testing different hypotheses
what to do to survive as long as possible)




## Supervised Learning

### Formalism




Let $\mathbf{X}=\{\mathfrak{X}_1,\dots,\mathfrak{X}_n\}$
be an input sample ("a database")
that consists of $n$ objects.



Most often we assume that each object $\mathfrak{X}_i$
is represented using $p$ numbers for some $p$.

We denote this fact as $\mathfrak{X}_i\in \mathbb{R}^p$
(it is *a $p$-dimensional real vector* or
*a sequence of $p$ numbers* or
*a point in a $p$-dimensional real space*
or *an element of a real $p$-space* etc.).

If we have "complex" objects on input,
we can always try representing them as **feature vectors** (e.g.,
come up with numeric attributes that best describe them in a task at hand).

{ BEGIN exercise }
Consider the following problems:

1. How would you represent a patient in a clinic?

2. How would you represent a car in an insurance company's database?

3. How would you represent a student in an university?
{ END exercise }





Of course, our setting is *abstract* in the sense that
there might be different realities *hidden* behind these symbols.

This is what maths is for -- creating *abstractions* or *models*
of complex entities/phenomena so that they can be much more easily manipulated
or understood.
This is very powerful -- spend a moment
contemplating how many real-world situations fit into our framework.



This also includes image/video data, e.g., a 1920×1080 pixel image
can be "unwound" to a "flat" vector of length 2,073,600.

(\*) There are some algorithms such as Multidimensional Scaling,
Locally Linear Embedding, IsoMap etc.
that can do that automagically.

. . .



In cases such as this we say that we deal with *structured (tabular) data*\
-- $\mathbf{X}$ can be written as an ($n\times p$)-matrix:
\[
\mathbf{X}=
\left[
\begin{array}{cccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} \\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} \\
\end{array}
\right]
\]
Mathematically, we denote this as $\mathbf{X}\in\mathbb{R}^{n\times d}$.





Remark.

: Structured data == think: Excel/Calc spreadsheets, SQL tables etc.








For an example, consider the famous Fisher's Iris flower dataset,
see `?iris` in R
and https://en.wikipedia.org/wiki/Iris_flower_data_set.

```{r show_iris}
X <- iris[1:6, 1:4] # first 6 rows and 4 columns
X         # or: print(X)
dim(X)    # gives n and p
dim(iris) # for the full dataset
```




$x_{i,j}\in\mathbb{R}$
represents the $j$-th feature of the $i$-th observation,
$j=1,\dots,p$, $i=1,\dots,n$.

For instance:

```{r show_iris_elem}
X[3, 2] # 3rd row, 2nd column
```





The third observation (data point, row in $\mathbf{X}$)
consists of items $(x_{3,1}, \dots, x_{3,p})$ that can be extracted by calling:

```{r show_iris_row}
X[3,]
as.numeric(X[3,]) # drops names
```

```{r show_iris_row_length}
length(X[3,])
```




Moreover, the second feature/variable/column
is comprised of
$(x_{1,2}, x_{2,2}, \dots, x_{n,2})$:

```{r show_iris_col}
X[,2]
length(X[,2])
```





We will sometimes use the following notation to emphasise that
the $\mathbf{X}$ matrix  consists of $n$ rows
or $p$ columns:

\[
\mathbf{X}=\left[
\begin{array}{c}
\mathbf{x}_{1,\cdot} \\
\mathbf{x}_{2,\cdot} \\
\vdots\\
\mathbf{x}_{n,\cdot} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
\mathbf{x}_{\cdot,1} &
\mathbf{x}_{\cdot,2} &
\cdots &
\mathbf{x}_{\cdot,p} \\
\end{array}
\right].
\]




Here, $\mathbf{x}_{i,\cdot}$ is a *row vector* of length $p$,
i.e., a $(1\times p)$-matrix:

\[
\mathbf{x}_{i,\cdot} = \left[
\begin{array}{cccc}
x_{i,1} &
x_{i,2} &
\cdots &
x_{i,p} \\
\end{array}
\right].
\]

Moreover, $\mathbf{x}_{\cdot,j}$ is a *column vector* of length $n$,
i.e., an $(n\times 1)$-matrix:

\[
\mathbf{x}_{\cdot,j} = \left[
\begin{array}{cccc}
x_{1,j} &
x_{2,j} &
\cdots &
x_{n,j} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{x}_{1,j} \\
{x}_{2,j} \\
\vdots\\
{x}_{n,j} \\
\end{array}
\right],
\]

where $\cdot^T$ denotes the *transpose* of a given matrix --
think of this as a kind of rotation; it allows us to introduce a set of
"vertically stacked" objects using a single inline formula.



### Desired Outputs




In supervised learning,
apart from the inputs we are also given the corresponding
reference/desired outputs.

The aim of supervised learning is to try to create an "algorithm" that,
given an input point, generates an output that is as *close* as possible
to the desired one. The given data sample will be used to "train" this "model".

Usually the reference outputs are encoded as individual  numbers (scalars)
or textual labels.

In other words, with each input $\mathbf{x}_{i,\cdot}$ we associate
the desired output $y_i$:

```{r show_iris_sample,echo=-1}
set.seed(123)
# in iris, iris[, 5] gives the Ys
iris[sample(nrow(iris), 3), ]  # three random rows
```





Hence, our dataset is $[\mathbf{X}\ \mathbf{y}]$ --
each object is represented as a row vector
$[\mathbf{x}_{i,\cdot}\ y_i]$, $i=1,\dots,n$:

\[
[\mathbf{X}\ \mathbf{y}]=
\left[
\begin{array}{ccccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} & y_1\\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} & y_2\\
\vdots & \vdots & \ddots & \vdots    & \vdots\\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} & y_n\\
\end{array}
\right],
\]

where:

\[
\mathbf{y} = \left[
\begin{array}{cccc}
y_{1} &
y_{2} &
\cdots &
y_{n} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{y}_{1} \\
{y}_{2} \\
\vdots\\
{y}_{n} \\
\end{array}
\right].
\]



### Types of Supervised Learning Problems




Depending on the type of the elements in $\mathbf{y}$
(the domain of $\mathbf{y}$),
supervised learning problems are usually
classified as:

- **regression** -- each $y_i$ is a real number

    e.g., $y_i=$ future market stock price
    with $\mathbf{x}_{i,\cdot}=$ prices from $p$ previous days

- **classification** -- each $y_i$ is a discrete label

    e.g., $y_i=$ healthy (0) or ill (1)
    with $\mathbf{x}_{i,\cdot}=$ a patient's health record


- **ordinal regression** (a.k.a. ordinal classification) -- each $y_i$ is a rank

    e.g., $y_i=$ rating of a product on the scale 1--5
    with $\mathbf{x}_{i,\cdot}=$ ratings of $p$ most similar products





{ BEGIN exercise }
Example Problems -- Discussion:

Which of the following are instances of classification problems? Which of them are regression tasks?

What kind of data should you gather in order to tackle them?

- Detect email spam
- Predict a market stock price
- Predict the likeability of a new ad
- Assess credit risk
- Detect tumour tissues in medical images
- Predict time-to-recovery of cancer patients
- Recognise smiling faces on photographs
- Detect unattended luggage in airport security camera footage
- Turn on emergency braking to avoid a collision with pedestrians
{ END exercise }




A single dataset can become an instance of many different ML problems.

Examples -- the `wine_quality` dataset:

```{r wines_load,echo=-1}
set.seed(123)
wine_quality <- read.csv("datasets/wine_quality_all.csv",
    comment.char="#")
wine_quality[1,]
```


`alcohol` is a numeric (quantitative) variable (see Figure {@fig:wines_regression} for a histogram depicting its empirical distribution):


```{r wines_regression,fig.cap="Quantitative (numeric) outputs lead to regression problems"}
summary(wine_quality$alcohol) # continuous variable
hist(wine_quality$alcohol, main="", col="white"); box()
```

`color` is a quantitative variable with two possible outcomes (see Figure {@fig:wines_binary} for a bar plot):



```{r wines_binary,fig.cap="Quantitative outputs lead to classification tasks"}
table(wine_quality$color) # binary variable
barplot(table(wine_quality$color), col="white", ylim=c(0, 6000))
```


Moreover, `response` is an ordinal variable, representing
a wine's rating as assigned by a wine expert
(see  Figure {@fig:wines_ordinal} for a barplot).
Note that although the ranks are represented with numbers,
we they are not continuous variables. Moreover,
these ranks are something more than just labels -- they are linearly
ordered, we know what's the smallest rank and whats the greatest one.

```{r wines_ordinal,fig.cap="Ordinal variables constitute ordinal regression tasks"}
table(wine_quality$response) # ordinal variable
barplot(table(wine_quality$response), col="white", ylim=c(0, 3000))
```


