<!--
kate: indent-width 4; word-wrap-column 74; default-dictionary en_AU
Copyright (C) 2020, Marek Gagolewski, https://www.gagolewski.com
This material is licensed under the Creative Commons BY-NC-ND 4.0 License.
-->

# Introduction {#chap:introduction}

<!-- I have no idea why setting this in the preamble is ignored: -->

{ LATEX \renewcommand{\mathbf}[1]{\mathbfup{#1}} }
{ LATEX \renewcommand{\boldsymbol}[1]{\mathbfit{#1}} }



<!-- TODO: citations


-->

An *algorithm* is a well-defined sequence of instructions that,
for a given sequence of input arguments,
yields some desired output.
In other words, it is a specific recipe for what we call
a *function* in mathematics.
Unfortunately, algorithm development is a tedious task. We need to be
super-precise about covering all the possible scenarios and modelling them
accurately. When we are to build a self-driving vehicle, we need to be
ready for an infinite
number of situations that can arise on the road.
When we are to build a system for diagnosing tumour on medical images,
there are so many decisions
to make of whether should we consider a particular area suspicious,
how does it relate to the neighbouring cells, etc.



## What is Machine Learning?


In *machine learning* (ML), we build and study computer algorithms
that make *predictions* or *decisions* but whose *all tiny bits* are not
manually programmed. Moreover, some algorithms might also be able
to *discover* new interesting facts about a problem instance at hand.

*Learning*, however, needs some material based upon which
new knowledge is to be developed.
In other words, we need *data*.



### Data Sources

Information can come from various sources, e.g., physical sensors,
files, databases, or (pseudo)random number generators.
It can take different forms, e.g., vectors, matrices
and other tensors, graphs, audio/video streams, text etc.
With the advent of the internet era, data have become ubiquitous.

{ BEGIN exercise }
Think of how much information you consume and generate when you
interact with your social media or news feeds every day.
{ END exercise }



Here are some application domains where machine learning has already proven
itself very useful:

* Financial services (banking, insurance, investment funds)

* Oil, gas, and other energy (solar, mining)

* Real estate

* Pharmaceuticals

* Advertising

* Transportation

* Retail

* Healthcare

* Food production




{ BEGIN exercise }
Think of different ways in which
these sectors could benefit from ML solutions.
{ END exercise }




To be frank, the above list was generated by duckduckgoing
the "biggest industries" query. That was a very easy task;
ML is already everywhere.
Basically, wherever we have data and there is a need to improve
some processes or discover new aspects about a problem domain,
there is a place for ML solutions.

Of course, it's not all about business revenue (luckily).
We can do a lot of great work for greater good;
with the increased availability of open data,
everyone can be a reporter, an engaged citizen that seeks for truth.
There are NGOs.
Finally, there are researchers (remember that the  main role of most
universities is still to spread the advancement of knowledge and not to make money!)
that need these methods to make new discoveries, e.g., in psychology,
sociology, agriculture, engineering, biotechnology, pharmacy, medicine, genetics,
you name it.

<!-- * Life science and medical data analysis -->
<!-- * Scientific simulations -->
<!-- * Genomics -->



### Main Types of Machine Learning Problems

Machine Learning problems include, but are not limited to:

* *Supervised learning* -- for every input point (e.g., a photo)
there is an associated desired output (e.g., whether it depicts a crosswalk
or how many cars can be seen on it);
in other words, there is a "teacher" ready to instruct us
about what is  the expected behaviour in a given context.


* *Semi-supervised learning* -- some inputs are labelled, other ones
are not (definitely a less laborious/cheaper scenario).



* *Unsupervised learning* -- inputs are unlabelled; the aim is to discover
the underlying structure in the data (e.g., automatically group customers
w.r.t. common behavioural patterns);
think about how much you learnt by just interacting with your
environment by wandering here and there.



* *Reinforcement learning* -- learn to act based on a
feedback given after the actual decision was made
such as "that was nice! well done! look at you go!"
(e.g., learn to play some video game by testing different hypotheses
what to do in order to prosper/survive as long as possible).



In the course of this book, we're going to take a deep dive into
the machine learning algorithms that stood the test of time
and that are still  used -- with modifications -- by researchers
and practitioners.
We will learn about how we can use them for improving different processes.
Importantly, we will also get to know their limitations.

Arranging the material so as to maximise its usefulness
(we'd like to empower ourselves with a set of tools that can get some work done),
lay solid groundwork for further studies and
develop the adjoining skills (the practice of data analysis certainly
requires us to learn programming and be able not only to read but also
understand mathematical
notation) is a multi-objective optimisation problem.
A problem involving many constraints though;
as this is an introductory course, we obviously cannot cover everything,
not yet. In particular, we will be learning maths and programming "from scratch",
our space-time is limited etc. We also want to enjoy our journey -- to have
some fun knowing that big projects never come into being in a day or seven.

Therefore, this time we will limit ourselves "just" to supervised and unsupervised
learning, as these are the most common  instances.




## Input Data, **X**

### Abstract Formalism


Let $\mathcal{X}=\{\boldsymbol{x}^{(1)},\dots,\boldsymbol{x}^{(n)}\}$
be an input sample ("a database")
that contains information on $n$ entities or phenomena (in their entirety
or at different points in time or space).
Most often we assume that each object in $\mathcal{X}$, say $\boldsymbol{x}^{(i)}$,
is represented by means of a sequence of $p$ real numbers (for some $p$).
We denote this fact as $\boldsymbol{x}^{(i)}\in \mathbb{R}^p$.



{ BEGIN remark }
Here, the "$\in$" symbol  stand for "*is in*" or "*is a member of*",
$\mathbb{R}$ denotes the set of real numbers (the one that features,
$0$, $-358745.2394$, $42$ and $\pi$, amongst uncountably many others)
and $\mathbb{R}^p$ is the set of real-valued sequences of length $p$
(i.e., $p$ such numbers considered at a time).
For instance, $\mathbb{R}^2$ includes pairs such as
$(1, 2)$, $(-6.4,0)$, and $(1/3, 10^{3})$.
More generally, we write $(x_1,\dots,x_p)\in\mathbb{R}^p$,
where $x_j\in\mathbb{R}$ is the $j$-th component of the sequence,
for any $j=1,\dots,p$.
{ END remark }

{ BEGIN remark }
To put emphasis on the fact that $(x_1,\dots,x_p)$ is not an "atomic"
entity, but a one that a collection of many *scalar* components,
we will use bold fonts for denoting whole sequences, e.g.,
write $\boldsymbol{x}=(x_1,\dots,x_p)$.
{ END remark }

{ BEGIN remark }
If $\boldsymbol{x}\in \mathbb{R}^p$, then we often say
that $\boldsymbol{x}$ is
*a sequence of $p$ numbers*,
*a (numeric) $p$-tuple*,
*a $p$-dimensional real vector*,
*a point in a $p$-dimensional real space*,
or *an element of a real $p$-space* etc.
We explain the basics of notation convention used throughout this book
(and in most math works in general) in Appendix \@ref(appendix:convention).
{ END remark }


At a first glance, such a representation might seem overly simplistic,
especially if we'd wish to store information on  "very complex" objects.
However, it turns out that in most cases expressing them as
*feature vectors* (i.e., establishing a set of numeric attributes
that best describe them in a task at hand) is not only natural
but also perfectly sufficient for achieving whatever we aim at.

<!--(\*) There are some algorithms such as Multidimensional Scaling,
Locally Linear Embedding, IsoMap etc.
that can do that automagically.-->


{ BEGIN exercise }
Consider the following problems:

* How would you represent a patient in a clinic
(for the purpose of conducting research in cardiology)?

* How would you represent a car in an insurance company's database
(for the purpose of determining how much a driver should pay annually
for the mandatory policy)?

* How would you represent a student in an university (for the purpose
of granting them scholarships)?

In each case, list a few numeric features that best describe the reality
of concern. Note that descriptive labels can always be *encoded* as numbers,
e.g., female = 1, male = 2.
{ END exercise }




The setting we've established is *abstract* in the sense that
there might be different realities *hidden* behind the symbols we use.
This is exactly what maths is for -- creating *abstractions* or *models*
of complex entities/phenomena so that they can be much more easily manipulated
or understood.



This is very powerful -- spend a moment
contemplating how many real-world situations fit into our framework.
In particular the last chapters are devoted to the less-obvious data domains,
for instance text or image/video data, e.g., a 1920×1080 pixel image
can be "unwound" to a "flat" sequence of length 2,073,600.



. . .

The whole dataset consisting of $n$ *observations* (samples, cases, records)
and $p$ *features* (variables, attributes, characteristics)
can be written as an ($n\times p$)-matrix:
\[
\mathbf{X}=
\left[
\begin{array}{cccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} \\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} \\
\end{array}
\right].
\]
Mathematically, we denote this as $\mathbf{X}\in\mathbb{R}^{n\times p}$
(read "$\mathbf{X}$ is a real-valued matrix with $n$ rows and $p$ columns").


{ BEGIN remark }
A *matrix* is an array of rectangular shape, with items
arranged in rows and columns.
In cases such as this we say that we deal with *structured (tabular) data*
(think: Excel/Calc spreadsheets, SQL tables etc.).
{ END remark }




### Concrete Example

Let's consider the Wine Quality dataset
(see Section \@ref(sec:wine-quality) for more details),
that stores information on
11 physicochemical features (amongst others -- see below)
of ca. 6500 Portuguese wines
from different wineries.


```{r wines-load,results='hide'}
wine_quality <- read.csv("datasets/wine_quality_all.csv",
    comment.char="#")
X <- wine_quality[, 1:11] # extract the first 11 columns
X <- as.matrix(X) # convert to numeric matrix
head(X) # display the first 6 rows, i.e., X[1:6,]
```

```{r wines-print1,echo=FALSE}
.X <- wine_quality[, 1:11]
.X <- .X[c(1,2,3,4,11)]
names(.X)[4] <- "..."
row.names(.X) <- sprintf("[%d,]", seq_len(nrow(.X)))
.X[4] <- "..."
print(head(.X))
```

We can get to now the dimensionality (shape) of $\mathbf{X}$ by calling:

```{r wines-print2}
dim(X)    # gives n and p, respectively
nrow(X)   # dim(X)[1] – 1st dimension – n
ncol(X)   # dim(X)[2] – 2nd dimension – p
```

Therefore, $\mathbf{X}\in\mathbb{R}^{`r nrow(X)`\times `r ncol(X)`}$,
i.e., $n=`r nrow(X)`$ and $p=`r ncol(X)`$.


{ BEGIN remark }
Obviously, displaying $\mathbf{X}$ in its entirety would take up a lot of
pages. Not only there are so many rows, also the number of columns makes
the matrix too "wide" -- mainly because
of the presence of the descriptive feature labels (column names).
Therefore, when printed out on the console, R will "wrap" the columns to
fit the page width.

```{r wines-print3}
head(X, 3) # the first 3 rows
```
{ END remark }

Let's forget about the column labels for the time being, as they
might be distracting use from fully appreciating the abstract
nature of the setting we're trying to develop.


```{r wines-print5}
dimnames(X) <- NULL # drop row and column names
head(X) # show the first 6 rows
```


An element of $\mathbf{X}$ located in the $i$-th row and the $j$-th column,
denoted $x_{i,j}\in\mathbb{R}$,
represents the $j$-th feature of the $i$-th observation,
$i=1,\dots,n$, $j=1,\dots,p$.

For instance, the 3rd wine's volatile acidity (expressed in grams of acetic acid per dm³), $x_{3,2}$, is:

```{r show-wines-elem}
X[3, 2] # 3rd row, 2nd column
```





The third observation (data point, row in $\mathbf{X}$, i.e.,
the 3rd wine in the dataset)
consists of items $(x_{3,1}, \dots, x_{3,p})$ that can be extracted by calling:

```{r show-wines-row}
X[3,]
```



Moreover, the second feature/variable/column (in our case: volatile acidity)
is comprised of
$(x_{1,2}, x_{2,2}, \dots, x_{n,2})$:

```{r show-wines-col}
X[,2]
```


{ BEGIN exercise }
Identify the precise location of the value `r X[3,2]` in all the above
examples.
{ END exercise }


We will sometimes use the following notation to emphasise that
the $\mathbf{X}$ matrix  consists of $n$ rows
or $p$ columns:

\[
\mathbf{X}=\left[
\begin{array}{c}
\mathbf{x}_{1,\cdot} \\
\mathbf{x}_{2,\cdot} \\
\vdots\\
\mathbf{x}_{n,\cdot} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
\mathbf{x}_{\cdot,1} &
\mathbf{x}_{\cdot,2} &
\cdots &
\mathbf{x}_{\cdot,p} \\
\end{array}
\right].
\]




Here, $\mathbf{x}_{i,\cdot}$ is a *row vector* of length $p$,
i.e., a $(1\times p)$-matrix:

\[
\mathbf{x}_{i,\cdot} = \left[
\begin{array}{cccc}
x_{i,1} &
x_{i,2} &
\cdots &
x_{i,p} \\
\end{array}
\right].
\]

Moreover, $\mathbf{x}_{\cdot,j}$ is a *column vector* of length $n$,
i.e., an $(n\times 1)$-matrix:

\[
\mathbf{x}_{\cdot,j} = \left[
\begin{array}{cccc}
x_{1,j} &
x_{2,j} &
\cdots &
x_{n,j} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{x}_{1,j} \\
{x}_{2,j} \\
\vdots\\
{x}_{n,j} \\
\end{array}
\right],
\]

where $\cdot^T$ denotes the *transpose* of a given matrix --
think of this as a kind of rotation; it allows us to introduce a set of
"vertically stacked" objects using a single inline formula.



{ BEGIN remark }
Details on how to get started with the R environment are provided
in Appendix \@ref(appendix:rintro).
A comprehensive introduction to vector and matrix algebra in R
as well as basic data frame wrangling is given
in Appendices \@ref(appendix:rvector), \@ref(appendix:rmatrix),
and \@ref(appendix:rdf).
From now on, we assume that we are all familiar with this material.
{ END remark }




## Unsupervised Learning



In *unsupervised learning* (learning without a teacher),
the input data vectors $\mathbf{x}_{1,\cdot},\dots,\mathbf{x}_{n,\cdot}$
(rows in $\mathbf{X}$) are all we have.
Our aim now is to discover the *underlying structure in the data*,
whatever that means.
For instance, Figure \@ref(fig:unsupervised) depicts a scatter plot
of two selected wine features.
What interesting information can we extract from this figure?
Perhaps not much, this is exactly why we need data mining methods!


```{r unsupervised,echo=FALSE,fig.cap="Unsupervised learning: \"But what it is exactly that I have to do here?\""}
plot(wine_quality[, "chlorides"], wine_quality[, "fixed.acidity"], col="#00000011", pch=16)
#plot(wine_quality[sam, "chlorides"], wine_quality[sam, "fixed.acidity"], col=wc[sam], pch=wp[sam])
```






### Dimensionality Reduction


Certain classes of unsupervised learning
problems are not only intellectually stimulating,
but practically useful at the same time.




In particular, in *dimensionality reduction* we seek a meaningful
*projection* of a high dimensional space (think: a matrix with many columns)
to a space of lower dimension.
For instance, if the output number of features is 2 (or maybe 3 as well),
we will be able to visualise a complex dataset on a standard scatter plot.


```{r princomp,echo=FALSE,fig.cap="Principal component analysis (a dimensionality reduction technique) applied on 5 columns of the `wine_quality` dataset."}
p <- princomp(scale(X[,c(1, 2, 4, 5, 11)]))
wc <- c("#44444422", "#ff000022")[as.numeric(factor(wine_quality$color, levels=c("white", "red")))]
wp <- c(17, 16)[as.numeric(factor(wine_quality$color, levels=c("white", "red")))]
set.seed(123)
sam <- sample(rev(1:nrow(X)))
plot(p$scores[sam,1], p$scores[sam,2], col=wc[sam], pch=wp[sam], xlab="First principal component", ylab="Second principal component")
legend("topright", c("white wine", "red wine"), pch=c(17,16), col=c("black", "red"), bg="white")
```

For instance, Figure \@ref(fig:princomp) depicts the two dimensions "revealed"
by the PCA method (Principal Component Analysis) -- a technique proposed
by the mathematician Karl Pearson (the Father of mathematical statistics)
at the beginning of the 20th century.

It's actually nice to see that the wines of different colours
(see `wine_quality[,"color"]`) form two quite distinct "blobs".
By analysing the results returned by PCA in more detail, we could
better understand the differences in physicochemical features of wines
of different colours (amongst others).

Once we go through this course,
we will be perfectly equipped for studying
such dimensionality reduction methods as:
multidimensional scaling (MDS), kernel PCA,
t-SNE, autoencoders (deep learning).
See, for example, [@esl] for more details.




### Anomaly Detection

Another family of unsupervised learning methods deals with
*anomaly detection*, where our task is to identify rare,
suspicious, ab-normal, novel or out-standing items.
For example, these can be cars driving on walkways in a park's security camera footage or fraudulent transactions identified in some financial service.



```{r anomaly-detection,echo=FALSE,fig.cap="Outliers can be thought of as anomalies of some sort"}
plot(p$scores[sam,1], p$scores[sam,2], col="#00000011", xlab="First principal component", ylab="Second principal component")
wh <- p$scores[,1]>10 | p$scores[,2]>4 # a "visual" outlier detector ;)
points(p$scores[wh,1], p$scores[wh,2],
    col=4, cex=3, pch=4)
legend("topright", c("outlier"), pch=c(4), col=c(4), bg="white")
```

Figure \@ref(fig:anomaly-detection) highlights a few *outliers* -- observations
that differ (in the sense the are much farther away) significantly from
the rest of the points. They might be due to some measurement or data input
error or just be an inherent part of the dataset. Hence, it might be
interesting to inspect them in more detail.
Section \@ref(sec:graphicalsummaryresidual) includes one of the
simplest (and most popular) definition of an outlier for unidimensional data.
More advanced methods, see, e.g., [@anomalydetection_survey],
are based upon the algorithms that we describe also in this book.
<!-- TODO: HDBSCAN? -->



### Clustering


Moreover, the aim of *clustering* (grouping or segmentation) is to automatically discover
some *naturally occurring*
subgroups in the data set.
For example, these may be customers having different shopping patterns
(such as "young parents", "students", and "elders")
or protein sequences that are to be grouped by their function.



```{r clustering-illustration,echo=FALSE,cache=TRUE,fig.cap="Two clusters identified by the K-means algorithm ($K=2$)"}
Y <- wine_quality[, c("chlorides", "fixed.acidity")]
km <- kmeans(p$scores[,1:2], centers=2, nstart=10)
wc <- c("#61D04F77", "#2297E677")[km$cluster]
wp <- c(17, 16)[km$cluster]
set.seed(123)
plot(p$scores[sam,1], p$scores[sam,2], col=wc[sam], pch=wp[sam], xlab="First principal component", ylab="Second principal component")
legend("topright", c("Cluster 1", "Cluster 2"),
col=c("#61D04F", "#2297E6"), pch=c(17,16), bg="white")
```


Figure \@ref(fig:clustering-illustration) gives the results of running the K-means
algorithm (see Chapter \@ref(chap:kmeans)) on the discussed dataset.
The two identified clusters are quite similar to
the wine colours, despite the fact that this information has not been used
during the method's operation.

We discuss the most seminal clustering techniques
in Chapters \@ref(chap:hclust) (hierarchical agglomerative algorithms)
and \@ref(chap:kmeans) (K-means method),
see also, e.g., [@jainsurvey], [@wierzchon_klopotek], [@esl].




## Supervised Learning


### Desired Outputs, **y**




In supervised learning, there is a teacher;
apart from the inputs (in the form of a matrix $\mathbf{X}$),
we are also given the corresponding reference/desired outputs
(a vector $\boldsymbol{y})$, which
are  usually encoded as individual  numbers (scalars)
or textual labels.

For instance, in the `wine_quality` dataset,
this can be the `response` column, which gives the wine rating on
the scale 0 (bad) to 10 (excellent).

```{r show-wines-Xy,results="hide"}
y <- wine_quality[, "response"] # extract the `response` column
Xy <- cbind(X, y) # join X and y (column-bind) - for printing
head(Xy)
```

```{r show-wines-Xy2,echo=FALSE}
.Xy <- wine_quality[,c(1,2,3,4,11,12)]
names(.Xy)[4] <- "..."
row.names(.Xy) <- sprintf("[%d,]", seq_len(nrow(.X)))
.Xy[4] <- "..."
print(head(.Xy))
```



More formally, with each input $\mathbf{x}_{i,\cdot}$ we associate
the desired output $y_i$, $i=1,\dots,n$.
Hence, our dataset is $[\mathbf{X}\ \mathbf{y}]$, with
the $i$-th row being
$[\mathbf{x}_{i,\cdot}\ y_i]$:

\[
[\mathbf{X}\ \mathbf{y}]=
\left[
\begin{array}{ccccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} & y_1\\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} & y_2\\
\vdots & \vdots & \ddots & \vdots    & \vdots\\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} & y_n\\
\end{array}
\right],
\]

where:

\[
\mathbf{y} = \left[
\begin{array}{cccc}
y_{1} &
y_{2} &
\cdots &
y_{n} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{y}_{1} \\
{y}_{2} \\
\vdots\\
{y}_{n} \\
\end{array}
\right].
\]


Where do we get the reference outputs from? Sometimes they are
the part of the dataset gathering process themselves.
In other cases, they require a lot of work and money -- in particular,
in order to judge the wine quality, someone had to engage a couple of experts
and ask them to drink 6500 wine bottles! This decent piece of research
certainly required a lot of sacrifices.

Amazon's Mechanical Turk or Google's image reCAPTCHA
(identify images with pedestrian crossings) are examples of services for
crowdsourcing of such $\mathbf{y}$s.




### Types of Supervised Learning Problems



The aim of supervised learning is to try to create an "algorithm" that,
given an input point, generates an output that is as *close* (with respect to
some useful metrics) as possible to the desired one.
The given data sample will be used to "train" this "model".




Depending on the type of the elements in $\mathbf{y}$
(the domain of $\mathbf{y}$),
supervised learning problems are usually
pigeonholed as:

* *regression* -- each $y_i$ is a real number,
    e.g., $y_i=$ future market stock price
    with $\mathbf{x}_{i,\cdot}=$ prices from $p$ previous days;

* *classification* -- each $y_i$ is a discrete label,
    e.g., $y_i=$ healthy (0) or ill (1)
    with $\mathbf{x}_{i,\cdot}=$ a patient's health record;


* *ordinal regression* (a.k.a. ordinal classification) -- each $y_i$ is
    a rank, e.g., $y_i=$ rating of a TV series on the scale 1--5
    with $\mathbf{x}_{i,\cdot}=$ ratings of $p$ most similar productions.




{ BEGIN exercise }
Which of the following are instances of classification problems? Which of them are regression tasks?

* Detect email spam

* Predict a market stock price

* Predict the likeability of a new ad

* Assess credit risk

* Detect tumour tissues in medical images

* Predict time-to-recovery of cancer patients

* Recognise smiling faces on photographs

* Detect unattended luggage in airport security camera footage

* Turn on emergency braking to avoid a collision with pedestrians

What kind of data should you gather in order to tackle them?
{ END exercise }


Once we build a supervised learning algorithm, its
outputs can be interpreted by a human being and used (critically)
for whatever  purpose (e.g., a GP can utilise it for diagnosing illnesses).
Their outputs can be also provided to other algorithms to take
immediate actions (e.g., buy those market shares asap!).

We are going to cover quite a decent number of classical approaches
to data classification and regression, respectively:

* K-nearest neighbour classifiers (Chapter \@ref(chap:knn)),
decision trees (Chapter \@ref(chap:trees)),
logistic regression (Chapter \@ref(chap:logistic)),
and its generalisation in form of feed-forward deep neural networks
(Chapter \@ref(chap:images));

* K-nearest neighbour regression (Chapter <!-- TODO -->**TODO**),
regression trees (Chapter <!-- TODO -->**TODO**) and,
linear models for regression (Chapter <!-- TODO -->**TODO**).



### One Dataset -- Many Problems

The title of this subsection might sound worrying, but
we need to be clear about one thing: in mathematics (and all
fields that use maths as a universal language),
we yearn for problems. If we have a problem, an interesting problem
in particular, it means we can occupy ourselves for a long time doing
something exciting and challenging. We just love what we do. It's not
merely our work, it's our hobby and passion as well.

Therefore, by saying that a single dataset can become an instance
of many different ML problems, we should be pretty pretty excited.

For example, in the `wine_quality` dataset we have the following columns:

```{r wines-colnames}
names(wine_quality) # names of the columns
```


. . .

Any  of the 11 physicochemical features can become the source of a
regression task (with $\mathbf{X}$ being all the remaining 10 -- or less),
because they are numeric (quantitative) variables.
For instance, Figure \@ref(fig:wines-regression) depicts
the empirical distribution (density) of the `alcohol` variable in form
of a histogram. We could be curious if, for example, any other
chemical component predicts high alcohol content.

Here are some basic summary statistics of this variable:


```{r wines-regression,fig.cap="Quantitative (numeric) outputs lead to regression problems"}
hist(wine_quality[,"alcohol"], main="", col="white"); box()
summary(wine_quality[,"alcohol"])
```



The outputs generated by `summary()` include:

* `Min.` -- sample minimum (the smallest value in the sample);

* `1st Qu.` -- 1st quartile (25th percentile, quantile of order 0.25),
being the observation $q$ such that 25\% values are not greater than $q$
and 75\% values are not smaller than $q$;

* `Median` -- median (2nd quartile, 50th percentile, quantile of order 0.5),
being the observation "in the middle of the sample";

* `Mean` -- arithmetic mean;

* `3rd Qu.` -- 3rd quartile (75th percentile, quantile of order 0.75);

* `Max.` --  sample maximum.



{ BEGIN remark }
The 1st, 2nd, and 3rd quartile of $(3, 2, 5, 4, 1)$ is
2, 3, 4, respectively (it's easiest to compute the quantiles by ordering
the input sequence). Computing quantiles in the case where they are
not defined in an unambiguous manner requires some (weighted) averaging,
see `?quantile`. For instance, most often we'll say that the median of
$(3, 2, 5, 6, 1, 4)$ is 3.5.
{ END remark }

. . .

On the other hand, `color` is a qualitative variable
with two possible outcomes (i.e., binary; see Figure \@ref(fig:wines-binary)
for a bar plot).
Therefore, it can constitute a classification task.

Quantitative data can be summarised with a contingency table:

```{r wines-binary,fig.cap="Qualitative (here: binary) outputs lead to classification tasks"}
barplot(table(wine_quality[,"color"]), col="white", ylim=c(0, 6000))
table(wine_quality[,"color"])
```


. . .

Moreover, `response` is an ordinal variable, representing
a wine's rating as assigned by a wine expert
(see  Figure \@ref(fig:wines-ordinal) for a barplot).
Note that although the ranks are represented with numbers,
we they are not continuous variables per se. However,
these ranks are "something more" than just labels -- they are linearly
ordered, we know what's the smallest rank and what's the greatest one,
ranks 3 and 7 are not just different: they are very different, etc.

```{r wines-ordinal,fig.cap="Ordinal variables constitute ordinal regression tasks"}
barplot(table(wine_quality[,"response"]), col="white", ylim=c(0, 3000))
table(wine_quality[,"response"])
```



{ LATEX \color{gray} }

. . .

**TODO** .....

Machine learning is not a cure for all our pains and itches!
Don't get the false impression that everything is in the data
and "will discover itself". We still need expert knowledge at various stages
of working with data: collection, preprocessing, inference, prediction,
communication etc.; e.g., which models/variables we choose, which constraint
we add etc.

Many "big" tasks are, and will be better of when modelled by human
(e.g., using partial differential equations), such as prediction of weather.
Machine learning could be a good addition to tune-up some of the
underlying parameters or as a way to smartly aggregate multiple models
into one forecast.



. . .

{ LATEX \normalcolor }





. . .

In the next chapter we will discuss our first family of clustering algorithms.


