# Simple Linear Regression  {#chap:regression-simple}

<!-- (C) 2020 Marek Gagolewski, https://www.gagolewski.com -->



{ LATEX \color{gray} }

**TODO** In this chapter, we will:

* ...

* ...


overfitting, generalisation



{ LATEX \normalcolor }






<!-- TODO: citations


-->



<!-- TODO



move some quality metrics from Ch.2 to Ch.1

what is high correlation? 0.6 is not! time to stop tolerating weak
associations in the social sciences, you can do better, get
more data or don't do what you do... :/

Social sciences is built on  sand

If your data is crap or your models are weak, admit it and give up.




different coefficients of correlation (Spearman, Kendall) → Chapter 2



-->



## Simple Regression

### Introduction


### Side Note: K-NN Regression


TODO: later



The K-Nearest Neighbour scheme is intuitively pleasing.

No wonder it has inspired a similar approach for solving a regression task.

In order to make a prediction for a new point $\mathbf{x}'$:

1. find the K-nearest neighbours of  $\mathbf{x}'$ amongst the points in the train set,
denoted $\mathbf{x}_{i_1,\cdot}, \dots, \mathbf{x}_{i_K,\cdot}$,
2. fetch the corresponding reference outputs $y_{i_1}, \dots, y_{i_K}$,
3. return their arithmetic mean as a result,
\[\hat{y}=\frac{1}{K} \sum_{j=1}^K y_{i_j}.\]








Recall our modelling of the Credit Rating ($Y$)
as a function of the average Credit Card Balance ($X$)
based on the `ISLR::Credit` dataset.



```{r knnreg1}
library("ISLR") # Credit dataset
Xc <- as.matrix(as.numeric(Credit$Balance[Credit$Balance>0]))
Yc <- as.matrix(as.numeric(Credit$Rating[Credit$Balance>0]))
```



```{r knnreg2,message=FALSE}
library("FNN") # knn.reg function
x <- as.matrix(seq(min(Xc), max(Xc), length.out=101))
y1  <- knn.reg(Xc, x, Yc, k=1)$pred
y5  <- knn.reg(Xc, x, Yc, k=5)$pred
y25 <- knn.reg(Xc, x, Yc, k=25)$pred
```


The three models are depicted in Figure \@ref(fig:knnreg3).
Again, the higher the $K$, the smoother the curve. On the other hand, for
small $K$ we adapt better to what's in a point's neighbourhood.

```{r knnreg3,message=FALSE,fig.cap="K-NN regression example"}
plot(Xc, Yc, col="#666666c0",
    xlab="Balance", ylab="Rating")
lines(x, y1,  col=2, lwd=3)
lines(x, y5,  col=3, lwd=3)
lines(x, y25, col=4, lwd=3)
legend("topleft", legend=c("K=1", "K=5", "K=25"),
    col=c(2, 3, 4), lwd=3, bg="white")
```




**Simple regression** is the easiest setting to start with -- let's assume
$p=1$, i.e., all inputs are 1-dimensional.
Denote $x_i=x_{i,1}$.

We will use it to build many intuitions, for example, it'll be easy
to illustrate all the concepts graphically.

```{r credit_scatter,fig.cap="A scatter plot of  Rating vs. Balance"}
library("ISLR") # Credit dataset
plot(Credit$Balance, Credit$Rating) # scatter plot
```


In what follows we will be modelling the Credit Rating ($Y$)
as a function of the average Credit Card Balance ($X$) in USD
for customers *with positive Balance only*.
It is because it is evident from Figure \@ref(fig:credit_scatter)
that some customers with zero balance obtained a credit rating
based on some external data source that we don't have access to in
our very setting.

```{r credit_XY}
X <- as.matrix(as.numeric(Credit$Balance[Credit$Balance>0]))
Y <- as.matrix(as.numeric(Credit$Rating[Credit$Balance>0]))
```

Figure \@ref(fig:credit_XY_plot) gives the updated scatter plot
with the zero-balance clients "taken care of".

```{r credit_XY_plot,fig.cap="A scatter plot of  Rating vs. Balance with clients of Balance=0 removed"}
plot(X, Y, xlab="X (Balance)", ylab="Y (Rating)")
```

Our aim is to construct a function $f$ that
**models** Rating as a function of Balance,
$f(X)=Y$.

We are equipped with $n=`r length(X)`$ reference (observed) Ratings
$\mathbf{y}=[y_1\ \cdots\ y_n]^T$
for particular Balances $\mathbf{x}=[x_1\ \cdots\ x_n]^T$.





Note the following naming conventions:

* Variable types:

    - $X$ -- independent/explanatory/predictor variable

    - $Y$ -- dependent/response/predicted variable

* Also note that:

    - $Y$ -- idealisation (any possible Rating)

    - $\mathbf{y}=[y_1\ \cdots\ y_n]^T$ -- values actually observed







The model will not be ideal, but it might be usable:

- We will be able to **predict** the rating of any new client.


    What should be the rating of a client with Balance of \$1500?

    What should be the rating of a client with Balance of \$2500?

- We will be able to **describe** (understand) this reality using a single mathematical formula
so as to infer that there is an association between $X$ and $Y$

    Think of "data compression" and laws of physics, e.g., $E=mc^2$.








(\*) Mathematically, we will assume that there is some "true" function that models the data
(true relationship between $Y$ and $X$),
but the observed outputs are subject to **additive error**:
\[Y=f(X)+\varepsilon.\]


$\varepsilon$ is a random term, classically we assume that
errors are independent of each other,
have expected value of $0$ (there is no systematic error = unbiased)
and that they follow a normal distribution.




(\*) We denote this as $\varepsilon\sim\mathcal{N}(0, \sigma)$
(read: random variable $\varepsilon$ follows a normal distribution
with expected value of $0$ and standard deviation of $\sigma$ for some $\sigma\ge 0$).

$\sigma$ controls the amount of noise (and hence, uncertainty).
Figure \@ref(fig:normal_distribs) gives the plot of the probability
distribution function (PDFs, densities)
of $\mathcal{N}(0, \sigma)$ for different $\sigma$s:



```{r normal_distribs, echo=FALSE, message=FALSE, fig.cap="Probability density functions of normal distributions with different standard deviations $\\sigma$."}
y <- seq(-3,3,length.out=101)
plot(y, dnorm(y, mean=0), type='l', ann=FALSE, ylim=c(0,dnorm(0,0,0.5)))
lines(y, dnorm(y, mean=0, sd=0.5), type='l', col=2, lty=2)
legend("topleft", lty=c(1,2), col=c(1,2), legend=expression(sigma*"="*1, sigma*"="*0.5), bg="white")
abline(v=0, lty=3, col="gray")
```



### Search Space and Objective




There are many different functions that can be **fitted** into
the observed $(\mathbf{x},\mathbf{y})$,
compare Figure \@ref(fig:credit_different_models).
Some of them are better than the other (with respect to
different aspects, such as fit quality, simplicity etc.).


```{r credit_different_models, echo=FALSE, message=FALSE,fig.cap="Different polynomial models fitted to data"}
plot(X, Y, col="#000000ff")

#library("mda")
#f <- mars(as.matrix(X), Y, degree=5)
#x <- seq(min(X), max(X), length.out=101)
#y <- predict(f, x)
#lines(x,y,col=4,lty=4,lwd=2)

f <- lm(Y~poly(X, 2))
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=2,lty=1,lwd=3)

f <- lm(Y~poly(X, 15))
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=3,lty=2,lwd=3)

lines(X[order(X)], Y[order(Y)],col=4,lty=3,lwd=3)

abline(lm(Y~X), col=5, lwd=3, lty=4)
```

Thus, we need a formal **model selection criterion**
that could enable as to tackle the model fitting
task on a computer.







Usually, we will be interested in a model
that minimises appropriately aggregated **residuals**
$f(x_i)-y_i$, i.e.,
**predicted outputs minus observed outputs**,
often denoted with $\hat{y}_i-y_i$,
for $i=1,\dots,n$.

```{r credit_residuals, echo=FALSE, message=FALSE,fig.cap="Residuals are defined as the differences between the predicted and observed outputs $\\hat{y}_i-y_i$"}
plot(X, Y, col=1, xlim=c(1500,2000), ylim=c(400, 1000))
f <- lm(Y~poly(X, 2))
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=2,lty=1,lwd=2)

Y2 <- predict(f, data.frame(X=X))

segments(X, Y, X, Y2, lty=2)
points(X, Y2, pch=7, col="red")


legend("bottomright",
    legend=c("observed output", "predicted output", "fitted model"),
    pch=c(1,7, NA), lty=c(NA, NA, 1), lwd=c(NA, NA, 2),
    col=c(1, "red", 2), bg="white")
```


In Figure \@ref(fig:credit_residuals), the residuals correspond to the
lengths of the dashed line segments -- they measure the discrepancy between
the outputs generated by the model (what we get)
and the true outputs (what we want).






Top choice: sum of squared residuals:
\[
\begin{array}{rl}
\mathrm{SSR}(f|\mathbf{x},\mathbf{y})
& = \left( f(x_1)-y_1 \right)^2 + \dots + \left( f(x_n)-y_n \right)^2 \\
& =
\displaystyle\sum_{i=1}^n \left( f(x_i)-y_i \right)^2
\end{array}
\]





Remark.

: The notation $\mathrm{SSR}(f|\mathbf{x},\mathbf{y})$ means that
it is the error measure
corresponding to the model $(f)$ *given* our data.\
We could've  denoted it  with $\mathrm{SSR}_{\mathbf{x},\mathbf{y}}(f)$
or even $\mathrm{SSR}(f)$ to emphasise that $\mathbf{x},\mathbf{y}$
are just fixed  values and we are not  interested
in changing them at all (they are "global variables").






We enjoy SSR because (amongst others):

- larger errors are penalised much more than smaller ones

    > (this can be considered a drawback as well)



- (\*\*) statistically speaking, this has a clear underlying interpretation

    > (assuming errors are normally distributed,
    finding a model minimising the SSR is equivalent
    to maximum likelihood estimation)

- the models minimising the SSR can often be found easily

    > (corresponding optimisation tasks have an analytic solution --
    studied already by Gauss in the late 18th century)


(\*\*) Other choices:

- regularised SSR, e.g., lasso or ridge regression (in the case of multiple input variables)
- sum or median of absolute values (robust regression)








Fitting a model to data can be written as an optimisation problem:

\[
\min_{f\in\mathcal{F}} \mathrm{SSR}(f|\mathbf{x},\mathbf{y}),
\]

i.e., find $f$ minimising the SSR **(seek "best" $f$)**\
amongst the set of admissible models $\mathcal{F}$.



Example $\mathcal{F}$s:

- $\mathcal{F}=\{\text{All possible functions of one variable}\}$  --  if there are no repeated
$x_i$'s, this corresponds to data *interpolation*; note that there
are many functions that give SSR of $0$.

- $\mathcal{F}=\{ x\mapsto x^2, x\mapsto \cos(x), x\mapsto \exp(2x+7)-9 \}$ -- obviously
an ad-hoc choice but you can easily choose the best amongst the 3 by computing 3 sums of squares.

- $\mathcal{F}=\{ x\mapsto a+bx\}$ -- the space of linear functions of one variable

- etc.


(e.g., $x\mapsto x^2$ is read "$x$ maps to $x^2$" and is
an elegant way to define an inline function $f$ such that $f(x)=x^2$)




## Simple Linear Regression

### Introduction




If the family of admissible models $\mathcal{F}$ consists only of all linear functions of one variable,
we deal with a **simple linear regression**.

Our problem becomes:


\[
\min_{a,b\in\mathbb{R}} \sum_{i=1}^n \left(
a+bx_i-y_i
\right)^2
\]


In other words, we seek best fitting line in terms of the squared residuals.

This is the **method of least squares**.

This is particularly nice, because our search space
is just $\mathbb{R}^2$ -- easy to handle both analytically and numerically.




```{r credit_different_lines_ssr,echo=FALSE,fig.height=4,fig.cap="Three simple linear models together with the corresponding SSRs"}
plot(X, Y, col="#00000044")
m1 <- lm(Y~X)
abline(m1, col=2,lwd=3, lty=1)
m2 <- lm(Y~X+0)
abline(m2, col=3,lwd=3, lty=2)
m3 <- L1pack::lad(Y~X)
abline(m3, col=4,lwd=3, lty=4)
legend("topleft",
    lty=c(1,2,4),
    col=c(2,3,4),
    legend=c(
        sprintf("y=%03.0f+%.3fx; SSR=%.0f",m1$coefficients[1],m1$coefficients[2], sum(m1$residuals^2)),
        sprintf("y=%03.0f+%.3fx; SSR=%.0f",0,                 m2$coefficients[1], sum(m2$residuals^2)),
        sprintf("y=%03.0f+%.3fx; SSR=%.0f",m3$coefficients[1],m3$coefficients[2], sum(m3$residuals^2))
    ), bg="white")
```


{ BEGIN exercise }
Which of lines in Figure \@ref(fig:credit_different_lines_ssr) is the least squares solution?
{ END exercise }


<!-- TODO:
heatmap of SSR(a,b) for a = [0,1], b=[0,300]
-->





### Solution in R




Let's fit the linear model minimising the SSR in R.
The `lm()` function (`l`inear `m`odels) has a convenient *formula*-based interface.


```{r credit_fit_lm}
f <- lm(Y~X)
```

In R, the expression "`Y~X`" denotes a formula, which we read as:
variable `Y` is a function of `X`.
Note that the dependent variable is on the left side of the formula.
Here, `X` and `Y` are two R numeric vectors of identical lengths.


Let's print the fitted model:

```{r credit_print_lm}
print(f)
```

Hence, the fitted model is:
\[
Y = f(X) = `r f$coefficient[1]`+`r f$coefficient[2]` X
\qquad (+ \varepsilon)
\]


Coefficient $a$ (intercept):

```{r credit_print_lm_2}
f$coefficient[1]
```

Coefficient $b$ (slope):

```{r credit_print_lm_1}
f$coefficient[2]
```

Plotting, see Figure \@ref(fig:credit_plot_lm):

```{r credit_plot_lm, fig.cap="Fitted regression line"}
plot(X, Y, col="#000000aa")
abline(f, col=2, lwd=3)
```


SSR:

```{r credit_ssr_lm}
sum(f$residuals^2)
sum((f$coefficient[1]+f$coefficient[2]*X-Y)^2) # equivalent
```

We can predict the model's output for yet-unobserved inputs by
writing:

```{r credit_predict_lm}
X_new <- c(1500, 2000, 2500) # example inputs
f$coefficient[1] + f$coefficient[2]*X_new
```


Note that linear models can also be fitted based on formulas that refer
to a data frame's columns. For example, let us wrap
both $\mathbf{x}$ and $\mathbf{y}$ inside a data frame:

```{r credit_XY_df}
XY <- data.frame(Balance=X, Rating=Y)
head(XY, 3)
```

By writing:

```{r credit_XY_df_lm}
f <- lm(Rating~Balance, data=XY)
```

now `Balance` and `Rating` refer to column names in the `XY` data frame,
and not the objects in R's "workspace".

Based on the above, we can make a prediction
using the `predict()` function"

```{r credit_XY_predict}
X_new <- data.frame(Balance=c(1500, 2000, 2500))
predict(f, X_new)
```

Interestingly:

```{r credit_XY_predict_extrapolate}
predict(f, data.frame(Balance=c(5000)))
```

This is more than the highest possible rating -- we have extrapolated
way beyond the observable data range.




Note that our $Y=a+bX$ model is **interpretable**
and **well-behaving**
(not all machine learning models will have this feature,
think: deep neural networks,
which we rather conceive as *black boxes*):

* we know that by increasing $X$ by a small amount,
$Y$ will also increase (positive correlation),

* the model is continuous -- small change in $X$
doesn't yield any drastic change in $Y$,

* we know what will happen if we increase or decrease $X$ by, say, $100$,

* the function is invertible -- if we want Rating of $500$,
we can compute the associated preferred Balance that should yield it
(provided that the model is valid).


### Analytic Solution


It may be shown (which we actually do below)
that the solution is:

\[
\left\{
\begin{array}{rl}
b  = & \dfrac{
n \displaystyle\sum_{i=1}^n x_i y_i - \displaystyle\sum_{i=1}^n  y_i \displaystyle\sum_{i=1}^n x_i
}{
n \displaystyle\sum_{i=1}^n x_i x_i -   \displaystyle\sum_{i=1}^n x_i\displaystyle\sum_{i=1}^n x_i
}\\
a = & \dfrac{1}{n}\displaystyle\sum_{i=1}^n  y_i - b  \dfrac{1}{n} \displaystyle\sum_{i=1}^n x_i  \\
\end{array}
\right.
\]

Which can be implemented in R as follows:

```{r lm_estimate_manual}
n <- length(X)
b <- (n*sum(X*Y)-sum(X)*sum(Y))/(n*sum(X*X)-sum(X)^2)
a <- mean(Y)-b*mean(X)
c(a, b) # the same as f$coefficients
```

### Derivation of the Solution (\*\*)




Remark.

: You can safely skip this part if you are yet to know
how to search for a minimum of a function of many variables
and what are partial derivatives.



Denote with:

\[
E(a,b)=\mathrm{SSR}(x\mapsto a+bx|\mathbf{x},\mathbf{y})
=\sum_{i=1}^n \left( a+bx_i - y_i \right) ^2.
\]

We seek the minimum of $E$ w.r.t. both $a,b$.



Theorem.

: If $E$ has a (local) minimum at $(a^*,b^*)$,
then its partial derivatives vanish therein,
i.e., $\partial E/\partial a(a^*, b^*) = 0$ and $\partial E/\partial b(a^*, b^*)=0$.





We have:

\[
E(a,b)   = \displaystyle\sum_{i=1}^n \left( a+bx_i - y_i \right) ^2.
\]

We need to compute the partial derivatives $\partial E/\partial a$ (derivative of $E$
w.r.t. variable $a$ -- all other terms treated as constants)
and $\partial E/\partial b$ (w.r.t. $b$).



Useful rules -- derivatives w.r.t. $a$ (denote $f'(a)=(f(a))'$):

- $(f(a)+g(a))'=f'(a)+g'(a)$ (derivative of sum is sum of derivatives)
- $(f(a) g(a))' = f'(a)g(a) + f(a)g'(a)$ (derivative of product)
- $(f(g(a)))' = f'(g(a)) g'(a)$ (chain rule)
- $(c)' = 0$ for any constant $c$ (expression not involving $a$)
- $(a^p)' = pa^{p-1}$ for any $p$
- in particular: $(c a^2+d)'=2ca$, $(ca)'=c$, $((ca+d)^2)'=2(ca+d)c$ (application of the above rules)


<!--\[
\begin{array}{rl}
E_a'(a,b)=&2\displaystyle\sum_{i=1}^n \left( a+bx_i - y_i \right) x_i \\
E_b'(a,b)=&2\displaystyle\sum_{i=1}^n \left( a+bx_i - y_i \right) \\
\end{array}
\]-->

We seek $a,b$ such that $\frac{\partial E}{\partial a}(a,b) = 0$
and $\frac{\partial E}{\partial b}(a,b)=0$.




\[
\left\{
\begin{array}{rl}
\frac{\partial E}{\partial a}(a,b)=&2\displaystyle\sum_{i=1}^n \left( a+bx_i - y_i \right) = 0\\
\frac{\partial E}{\partial b}(a,b)=&2\displaystyle\sum_{i=1}^n \left( a+bx_i - y_i \right) x_i = 0 \\
\end{array}
\right.
\]

This is a system of 2 linear equations. Easy.

Rearranging like back in the school days:

\[
\left\{
\begin{array}{rl}
b \displaystyle\sum_{i=1}^n x_i+ a n = & \displaystyle\sum_{i=1}^n  y_i  \\
b \displaystyle\sum_{i=1}^n x_i x_i + a \displaystyle\sum_{i=1}^n x_i = & \displaystyle\sum_{i=1}^n x_i y_i \\
\end{array}
\right.
\]




It is left as an exercise to show that the solution is:

\[
\left\{
\begin{array}{rl}
b^*  = & \dfrac{
n \displaystyle\sum_{i=1}^n x_i y_i - \displaystyle\sum_{i=1}^n  y_i \displaystyle\sum_{i=1}^n x_i
}{
n \displaystyle\sum_{i=1}^n x_i x_i -   \displaystyle\sum_{i=1}^n x_i\displaystyle\sum_{i=1}^n x_i
}\\
a^* = & \dfrac{1}{n}\displaystyle\sum_{i=1}^n  y_i - b^*  \dfrac{1}{n} \displaystyle\sum_{i=1}^n x_i  \\
\end{array}
\right.
\]



(we should additionally perform the second derivative test
to assure that this is the minimum of $E$ -- which is exactly the case though)

<!-- TODO rewrite nicely with means and correlation coefficients etc.

Pearson's r introduced in the next chapter though

(a <- cor(x,y)*sd(y)/sd(x))
(b <- mean(y)-a*mean(x))
-->








(\*\*) In the next chapter, we will introduce the notion of Pearson's
linear coefficient, $r$ (see `cor()` in R). It might be shown that
$a$ and $b$ can also be rewritten as:

```{r ab_coef_cor}
(b <- cor(X,Y)*sd(Y)/sd(X))
(a <- mean(Y)-b*mean(X))
```

<!-- TODO show optim() -->



## Exercises

### The Anscombe Quartet

Here is a famous illustrative example proposed by
the statistician Francis Anscombe in the early 1970s.

```{r anscombe}
print(anscombe) # `anscombe` is a built-in object
```

What we see above is a single data frame
that encodes four separate datasets:
`anscombe$x1` and `anscombe$y1` define the first pair of variables,
`anscombe$x2` and `anscombe$y2` define the second pair and so forth.





{ BEGIN exercise }
Split the above data (manually) into four data frames
`ans1`, ..., `ans4` with columns `x` and `y`.

For example, `ans1` should look like:

```{r anscombe1,echo=-(1:4)}
ans1 <- data.frame(x=anscombe$x1, y=anscombe$y1)
ans2 <- data.frame(x=anscombe$x2, y=anscombe$y2)
ans3 <- data.frame(x=anscombe$x3, y=anscombe$y3)
ans4 <- data.frame(x=anscombe$x4, y=anscombe$y4)
print(ans1)
```
{ END exercise }





{ BEGIN solution }

```{r anscombe1apply, eval=FALSE}
<<anscombe1>>
```

{ END solution }



{ BEGIN exercise }
Compute the mean of each `x` and `y` variable.
{ END exercise }

{ BEGIN solution }

```{r anscombe_mean}
mean(ans1$x) # individual column
mean(ans1$y) # individual column
sapply(ans2, mean) # all columns in ans2
sapply(anscombe, mean) # all columns in the full anscombe dataset
```


*Comment: This is really interesting, all the means of `x` columns
as well as the means of `y`s are almost identical.*
{ END solution }




{ BEGIN exercise }
Compute the standard deviation of each `x` and `y` variable.
{ END exercise }

{ BEGIN solution }

The solution is similar to the previous one, just replace `mean` with `sd`.
Here, to learn something new, we will use the `knitr::kable()` function
that pretty-prints a given matrix or data frame:

```{r anscombe_sd}
results <- sapply(anscombe, sd)
knitr::kable(results, col.names="standard deviation")
```


*Comment: This is even more interesting, because the numbers agree up to 2 decimal digits.*

{ END solution }

<!--
TODO: cor() covered later
Compute the Pearson linear correlation coefficient
for each of the four pairs of `x` and `y`.

To recall, this can be done with the `cor()` function.
What is the Pearson $r$ coefficient is explained in Lecture 3.

```{r}
cor(ans1$x, ans1$y)
cor(ans2$x, ans2$y)
cor(ans3$x, ans3$y)
cor(ans4$x, ans4$y)
```


*Comment: We get a feeling that we're being tricked by Anscombe...
All the variables are highly correlated, and the correlation
coefficients are more or less the same.*
-->



{ BEGIN exercise }
Fit a simple linear regression model for each data set.
Draw the scatter plots again (`plot()`)
and add the regression lines (`lines()` or `abline()`).
{ END exercise }

{ BEGIN solution }

To recall, this can be done with the `lm()` function
explained in Lecture 2.

At this point we should already have become lazy -- the tasks are very
repetitious. Let's automate them by writing a single function
that does all the above for any data set:

```{r anscombe_fit}
fit_models <- function(ans) {
    # ans is a data frame with columns x and y
    f <- lm(y~x, data=ans) # fit linear model
    print(f$coefficients) # estimated coefficients
    plot(ans$x, ans$y) # scatter plot
    abline(f, col="red") # regression line
    return(f)
}
```

Now we can apply it on the four particular examples.

```{r anscombe_fit_apply,fig.height=6,fig.cap="Fitted regression lines for the Anscombe quartet"}
par(mfrow=c(2, 2)) # four plots on 1 figure (2x2 grid)
f1 <- fit_models(ans1)
f2 <- fit_models(ans2)
f3 <- fit_models(ans3)
f4 <- fit_models(ans4)
```

*Comment: All the estimated models are virtually the same,
the regression lines are
$y=0.5x+3$, compare Figure \@ref(fig:anscombe_fit_apply).*

{ END solution }





{ BEGIN exercise }
Create scatter plots of the residuals (predicted $\hat{y}_i$ minus
true $y_i$) as a function of the predicted $\hat{y}_i=f(x_i)$ for every
$i=1,\dots,11$.
{ END exercise }

{ BEGIN solution }

To recall, the model predictions can be generated by (amongst others)
calling the `predict()` function.


```{r anscombe_resid}
y_pred1 <- f1$fitted.values # predict(f1, ans1)
y_pred2 <- f2$fitted.values # predict(f2, ans2)
y_pred3 <- f3$fitted.values # predict(f3, ans3)
y_pred4 <- f4$fitted.values # predict(f4, ans4)
```

Plots of residuals as a function of the predicted (fitted) values
are given in Figure \@ref(fig:anscombe_resid_plot).

```{r anscombe_resid_plot,fig.height=6,fig.cap="Residuals vs. fitted values for the regression lines fitted to the Anscombe quartet"}
par(mfrow=c(2, 2)) # four plots on 1 figure (2x2 grid)
plot(y_pred1, y_pred1-ans1$y)
plot(y_pred2, y_pred2-ans2$y)
plot(y_pred3, y_pred3-ans3$y)
plot(y_pred4, y_pred4-ans4$y)
```

*Comment: Ideally, the residuals shouldn't be correlated
with the predicted values -- they should "oscillate" randomly
around 0.     This is only the case of the first dataset.
All the other cases are "alarming" in the sense that
they suggest that the obtained models are "suspicious"
(perhaps data cleansing is needed or a linear model is not at all appropriate).*

{ END solution }



{ BEGIN exercise }
Draw conclusions (in your own words).
{ END exercise }

{ BEGIN solution }
We're being taught a lesson here: don't perform data analysis tasks
automatically, don't look at bare numbers only, visualise your data first!
{ END solution }




{ BEGIN exercise }
Read more about Anscombe's quartet at https://en.wikipedia.org/wiki/Anscombe%27s_quartet
{ END exercise }




### Median House Value in Boston


The famous `Boston` dataset from the `MASS` package
records the historical (in the 1970s) median house value
(`medv` column, in 1000s USD) for 506 suburbs around
Boston, MA, USA.

You can access the dataset by calling:

```{r Boston_load1,message=FALSE}
# call install.packages("MASS") first (only once)
library("MASS")
head(Boston, 3)
```

Read the description of each of the 14 columns in the dataset's manual,
see `?Boston`.

{ BEGIN exercise }
Fit a simple linear model of $Y=$ `medv` as a function of $X=$  `lstat`.
{ END exercise }

{ BEGIN exercise }
Fit a quadratic polynomial model $(aX^2+bX+c)$ for the same pair of variables.
{ END exercise }

{ BEGIN exercise }
Fit a 5-degree polynomial model $(aX^5+bX^4+cX^3+dX^2+eX+f)$ for the same pair of variables.
{ END exercise }

{ BEGIN exercise }
Draw the scatter plot of the two variables and add the fitted regression
curves (all three on a single plot, use different colours).
{ END exercise }


{ BEGIN exercise }
Compute RMSE, MAE and $R^2$ of each model. Provide the interpretations of the
obtained values.
{ END exercise }

{ BEGIN exercise }
For each model, draw the plot of the residuals ($\hat{y}_i-y_i$)
as a function of the predicted outputs ($\hat{y}_i$).
Describe these plots in your own words.
{ END exercise }

{ BEGIN exercise }
Predict the `medv` values for `lstat` of 0, 25, 50 and 75
using all the models. Compare and discuss the results.
Which of the predictions seem trustworthy?
{ END exercise }




## Outro


In supervised learning, with each input point,
there's an associated reference output value.

Learning a model = constructing a function that approximates
(minimising some error measure) the given data.

Regression = the output variable $Y$ is continuous.

We studied linear models with a single independent variable based on
the least squares (SSR) fit.

In the next part we will extend this setting to the case
of many variables, i.e., $p>1$, called multiple regression.





{ LATEX \color{gray} }

. . .

**TODO** .....


. . .

{ LATEX \normalcolor }




Recommended further reading:  [@islr: Chapters 1, 2 and 3]

Other: [@esl: Chapter 1, Sections 3.2 and 3.3]

Next Chapter....


